### 2025-10-17 docs/Homeownership_BankOfMomAndDad.tex
Re stud (#20)
* some shortening

* old latex is corruped, work on this one in the meantime

* minor changes

* Make error bars thicker

* Better figures

* Small fix

* Update text

* FEDSstuff

* Taking on comments

---------

Co-authored-by: Eirik Eylands Brandsaas <eirik.e.brandsaas@frb.gov>

This paper quantifies the importance of parental transfers for young adults' homeownership and investigates how policies lowering homeownership barriers interact with parental transfers. A growing literature emphasizes that families matter for economic outcomes \citep{Doepke2016a}. In the context of homeownership, previous empirical studies highlight that parental transfer receipt is associated with an increase in the probability of becoming a homeowner, mainly by relaxing credit constraints \citep[see e.g.,][]{Lee2018,Blickle2019,wold2024housing}. Additionally, parental transfers help households maintain homeownership, a mechanism that has received less attention \citep[see e.g.,][]{bond2021role}. However, while these studies show that transfer receipt increases the probability of renters becoming homeowners, they do not establish the overall importance of parental transfers. Households may anticipate future transfers and reduce their own saving, so that the observed boost in homeownership conditional on receiving a transfer overstates the aggregate effect of transfers on homeownership.Furthermore, this is the first paper quantifying the importance of parental transfers as a determinant of homeownership; the existing literature has primarily focused on family formation and credit constraints \citep[see e.g.,][]{Chang2024,Paz-Pardo2019,Mabille2020}.

@@

I then study how illiquidity affects strategic behavior in the absence of commitment. As an illustration, consider a child who buys a house and brings no liquid wealth to the next period. If the parent does not transfer in the next period, the child will either face very low consumption—implying a high marginal value of wealth—or be forced to liquidate the house and incur sales costs. The parent dislikes both outcomes and may find it optimal to transfer enough to keep the child in the home. More precisely, children with wealthy parents use the sales costs as a commitment device to maintain low liquid wealth in the next period, raising their next-period marginal utility of wealth and thereby strengthening the parent’s transfer motive. As a result, 22\% of young households prefer illiquid housing over liquid housing. That parental wealth generates a preference for illiquidity is a new theoretical mechanism---distinct from existing explanations such as temptation or time inconsistency \citep[e.g.,][]{attanasio2024temptation,laibson1997golden}.

@@

	 \caption*{\begin{footnotesize} \textit{Notes:} The left panel uses the transfers, gifts, and inheritances variable, which is available in all waves of the PSID, while the right panel uses the parental-transfer data only available in the 2013 transfer supplement. See main text for a discussion. Vertical dotted and dashed lines denote the median and means, respectively. All estimates are weighted and omit the top 5\% of transfer sizes. \end{footnotesize}}
\end{figure}

@@

\subsubsection{Transfers and the Transition into Homeownership}
Since many papers have already estimated the relationship between parental transfers and transitions into homeownership (e.g., \citealp{wold2024housing}; \citealp{Blickle2019}; \citealp{benetton2022dynastic}; \citealp{Lee2018}), I replicate these regressions to highlight the association between transfer receipt and rent-to-own transitions, but relegate the full results to Appendix~\ref{app:rent_to_own}. Receiving a parental transfer above \$500 is associated with a 1.2 percentage point increase in the probability of becoming a homeowner (not statistically significant), after controlling for household characteristics. Receiving large transfers above \$10,000 is associated with a 7.9 percentage point increase in the probability of becoming a homeowner ($p<0.1$), even after including household fixed effects. These are sizable effects relative to the baseline rent-to-own transition probability of about 18\% during the sample period.

@@

\subsection{Taking Stock of the Data}
Taking stock, parental transfers have become more important for young homeowners over the past two decades. About 20\% of young households receive transfers from their parents in a given year, and recipients have wealthier parents. Recipients are more likely to transition from renting to owning, after conditioning on a rich set of covariates. Parental wealth is also positively associated with maintaining homeownership. Households with richer parents are less likely to fall behind on mortgages and to downsize during unemployment spells. These results suggest that homeowners with wealthy parents find the illiquidity of homeownership less problematic.

Throughout the empirical section, I do not treat transfer receipt or parental wealth as exogenous, and the estimates document associations rather than causal effects. For example, many transfers may be made precisely because the renter is buying a home. This motivates the structural model I describe in the next section, which provides a disciplined framework to interpret mechanisms and conduct counterfactuals.

@@

\textit{Housing:} Households can obtain housing services by renting or owning. They can rent housing of size $h_r$ or own houses of size $h_o$, with one size available for each tenure choice for tractability. The unit price of housing is $p$, and $q$ denotes the rent-to-price ratio. I assume that prices are constant for tractability, but I model aggregate house price risk as in \cite{Corbae2015} in Appendix \ref{sec:robust_pricerisk} and find that the quantitative results are almost unchanged. Homeowners incur proportional maintenance and depreciation costs $\delta$. Illiquidity is captured by proportional adjustment and moving costs on owner-occupied housing, as in \cite{Yang2009}:
\begin{equation}\label{eq:transcost}
adj(h_{a+1},h_a) = \begin{cases}
m_b p h_{a+1} & \text{if } h_{a}=h_r\mathbin{\&} h_{a+1}= h_o, \\
m_s p h_a 	& \text{if } h_a = h_o\mathbin{\&}h_{a+1}=h_r, \\
0 & \text{if } h_{a+1} = h_a,
\end{cases}
\end{equation} where $m_s$ and $m_b$ denote selling and buying costs, respectively. A household enters the period with housing $h_a$, while $h_{a+1}$ is the house chosen in this period.

@@

\textit{Housing Parameters:} I set the rental rate $q=0.10$ \citep{Davis2008}, housing depreciation $\delta=0.05$ \citep{Harding2007}, and sales cost $m_s=0.075$ and purchase cost $m_b=0.02$ \citep{Yang2009}. I set the price level to $p=\$232,918/h_o$ to match the average market value of owner-occupied houses for households aged 25 to 44 in my sample. The size $h_o$ is estimated internally.

@@

While the model and the PSID have the same two-year frequency, households were asked about transfers in the last calendar year. In my cleaned sample, the annual transfer rate is 24\%. Thus, the two-year transfer rate could range between 24\% and 48\%. To be very conservative, I assume that the two-year transfer rate equals the one-year transfer rate. My estimate of altruism $\eta$ should be viewed as a lower bound.\footnote{I have experimented with targeted higher transfer rates. This requires stronger altruism, which, in turn, increases homeownership motives, lowering the estimated utility premium $\chi$. Overall, the results are qualitatively similar to my conservative estimate, which understates the role of altruistic transfers.}

@@

\subsubsection{Household-Level Responses to Income Shocks by Parental Wealth}
As a final external validation, I repeat the exercise in Section \ref{sec:eventstudy} (see Figure \ref{fig:housinggrowthrates}), where we saw that households with wealthy parents do not downsize during unemployment spells. Since the model lacks an explicit unemployment state, households are considered unemployed if their productivity drops to the lowest level for a single period. Otherwise, I replicate the exercise exactly. Figure \ref{fig:housinggrowthrates_model} shows that the model matches both the qualitative and quantitative patterns from the empirical data relatively well: Housing consumption falls for households without wealthy parents during ``unemployment,'' while it does not decline for those with rich parents.

@@

\textit{Reducing Purchase Costs:}
Closing-cost programs and first-time-buyer relief---e.g., England’s 2017 stamp duty cut for first-time buyers---directly lower entry costs. I lower the purchase cost ($m_b$) by 2 percentage points, equivalent to a grant of about \$6,000. While homeownership rises, the gains are concentrated among households with wealthy parents, amplifying the parental wealth gradient. Lowering purchase costs has little effect for renters with non-wealthy parents, who remain constrained by LTV limits and sales costs. In contrast, for households with wealthy parents, lower purchase costs reduce the effective price—the main remaining barrier to ownership. This suggests that grants or credits for first-time buyers primarily benefit households with richer parents.

@@

\begin{table}[tb]
	\center\singlespacing	
	\begin{threeparttable}[tb]
		\caption{Altruistic Transfers Decrease the MPC}
		\input{../tabfig/postest/MPCs.tex}
	\end{threeparttable}
	{\begin{footnotesize}\begin{flushleft}\vspace{-0.1in}
		\textit{Notes:} The sample includes all households aged 25-44 in the stationary distribution. Following \cite{Kaplan2014}, households are liquidity constrained if they have negative liquid wealth $(b_k)$ exceeding 16.5\% of annual earnings. MPCs are calculated from a \$5,000 positive shock. ``Future transfers only'' MPCs use the baseline (altruistic) policy functions but set the contemporaneous transfer at the shock date to zero, while leaving expectations about future transfers unchanged. ``Current transfers only'' MPCs combine the parental policy function from the baseline with children’s policy functions solved without altruism; children receive the contemporaneous transfer but behave as if there are no future transfers. \end{flushleft}\end{footnotesize}}			
\end{table}

These results have several implications for the MPC literature. First, model-implied MPCs may be overstated when intra-family transfers are omitted, particularly for households with wealthy parents. Second, a common strategy to match high MPCs is to introduce households with low target wealth—e.g., through impatience \citep{aguiar2024hand}—but such behavior is observationally similar to that of households expecting altruistic support. Third, models that generate high MPCs by introducing illiquid assets with large return premia \citep{kaplan2022marginal} may overlook the fact that homeowners with wealthy parents have access to informal insurance, which lowers their effective MPC.

@@

Several simplifying assumptions keep the model tractable. The housing block could be enriched with multiple house sizes, refinancing frictions, and full general equilibrium to capture quality margins and price feedbacks. The dynasty is modeled as a single parent–child pair; in reality, resources are pooled across parents (including remarried/cohabiting partners) and allocated across multiple children. Pooling can raise transfer capacity, while sibling dilution spreads support, making the net strength of the transfer motive ambiguous and heterogeneous across families. Future work could also study whether illiquidity mitigates intrafamily commitment problems. Finally, the results on illiquidity preferences speak to the high-MPC literature \citep[e.g.,][]{kaplan2022marginal}: models that ignore parental transfers likely overstate average MPCs among liquidity-constrained households, some of whom rely on family support to smooth shocks.

>>>>

This paper quantifies the importance of parental transfers for young adults' homeownership and investigates how policies lowering homeownership barriers interact with parental transfers. Previous empirical studies show that receiving a transfer increases the probability of becoming a homeowner, mainly by relaxing credit constraints \citep[see e.g.,][]{Lee2018,Blickle2019,wold2024housing}, and that transfers help households maintain ownership \citep[e.g.,][]{bond2021role}. Yet these estimates capture only the direct effect of receiving a transfer, holding expectations and savings behavior fixed. In reality, households anticipate future support and adjust their saving and tenure decisions accordingly. Observed correlations between transfers and homeownership may thus overstate the aggregate role of transfers. I address this limitation by building a structural life-cycle model in which expectations of parental transfers, savings, and homeownership are jointly determined. Finally, this is the first paper quantifying the importance of parental transfers as a determinant of homeownership, whereas the existing literature has primarily focused on family formation and credit constraints \citep[see e.g.,][]{Chang2024,Paz-Pardo2019,Mabille2020}.

@@

I then study how illiquidity affects strategic behavior in the absence of commitment. As an illustration, consider a child who buys a house and brings no liquid wealth to the next period. If the parent does not transfer in the next period, the child will either face very low consumption---implying a high marginal value of wealth---or be forced to liquidate the house and incur sales costs. The parent dislikes both outcomes and may find it optimal to transfer enough to keep the child in the home. More precisely, children with wealthy parents use the sales costs as a commitment device to maintain low liquid wealth in the next period, raising their next-period marginal utility of wealth and thereby strengthening the parent’s transfer motive. As a result, 22\% of young households prefer illiquid housing over liquid housing. That parental wealth generates a preference for illiquidity is a new theoretical mechanism---distinct from existing explanations such as temptation or time inconsistency \citep[e.g.,][]{attanasio2024temptation,laibson1997golden}.

@@

	 \caption*{\begin{footnotesize} \textit{Notes:} The left panel uses the parental-transfer data only available in the 2013 transfer supplement, while the right panel uses the transfers, gifts, and inheritances variable, which is available in all waves of the PSID. See main text for a discussion. Vertical dotted and dashed lines denote the median and means, respectively. All estimates are weighted and omit the top 5\% of transfer sizes. \end{footnotesize}}
\end{figure}

@@

\subsubsection{Transfers and the Transition into Homeownership}
Since many papers have already estimated the relationship between parental transfers and transitions into homeownership (e.g., \citealp{wold2024housing}; \citealp{Blickle2019}; \citealp{benetton2022dynastic}; \citealp{Lee2018}; \citealp{kurz2004labour}; \citealp{spilerman2012parental}), I replicate these regressions in my sample to highlight the association between transfer receipt and rent-to-own transitions, but relegate the full results to Appendix~\ref{app:rent_to_own}. Receiving a parental transfer above \$500 is associated with a 1.2 percentage point increase in the probability of becoming a homeowner (not statistically significant), after controlling for household characteristics. Receiving large transfers above \$10,000 is associated with a 7.9 percentage point increase in the probability of becoming a homeowner ($p<0.1$), even after including household fixed effects. These are sizable effects relative to the baseline rent-to-own transition probability of about 18\% during the sample period.

@@

\subsection{Taking Stock of the Data}
Taking stock, parental transfers have become more important for young homeowners over the past two decades. About 20\% of young households receive transfers from their parents in a given year, and recipients have wealthier parents. Recipients are more likely to transition from renting to owning, after conditioning on a rich set of covariates. Parental wealth is also positively associated with maintaining homeownership: Households with richer parents are less likely to fall behind on mortgages and to downsize during unemployment spells, suggesting that homeowners with wealthy parents find the illiquidity of homeownership less problematic.

That said, the infrequency of homeownership transitions and small sample size limit statistical precision. Moreover, I do not treat transfer receipt or parental wealth as exogenous throughout the empirical section---the estimates document associations rather than causal effects. For example, many transfers may be made precisely because the renter is buying a home. Nonetheless, the patterns consistently point to strong associations between parental resources and both homeownership entry and persistence. This motivates the structural model I describe in the next section, which provides a disciplined framework to interpret mechanisms and conduct counterfactuals

@@

\textit{Housing:} Households can obtain housing services by renting or owning. They can rent housing of size $h_r$ or own houses of size $h_o$, with one size available for each tenure choice for tractability. The unit price of housing is $p$, and $q$ denotes the rent-to-price ratio. I assume that prices are constant for tractability, but I model aggregate house price risk as in \cite{Corbae2015} in Appendix \ref{sec:robust_pricerisk} and find that the quantitative results are almost unchanged. Homeowners must pay proportional maintenance and depreciation costs $\delta$. Illiquidity is captured by proportional adjustment and moving costs on owner-occupied housing, as in \cite{Yang2009}:
\begin{equation}\label{eq:transcost}
adj(h_{a+1},h_a) = \begin{cases}
m_b p h_{a+1} & \text{if } h_{a}=h_r\mathbin{\&} h_{a+1}= h_o, \\
m_s p h_a 	& \text{if } h_a = h_o\mathbin{\&}h_{a+1}=h_r, \\
0 & \text{if } h_{a+1} = h_a,
\end{cases}
\end{equation} where $m_s$ and $m_b$ denote selling and buying costs, respectively. A household enters the period with housing $h_a$, while $h_{a+1}$ is the house chosen in this period.

@@

\textit{Housing Parameters:} I set the rental rate $q=0.10$ \citep{Davis2008}, housing depreciation $\delta=0.05$ \citep{Harding2007}, and sales cost $m_s=0.075$ and purchase cost $m_b=0.02$ \citep{Yang2009}. I set the per-unit price level to $p=\$232,918/h_o$ to match the average market value of owner-occupied houses for households aged 25 to 44 in my sample. The size $h_o$ is estimated internally.

@@

While the model and the PSID have the same two-year frequency, the PSID asks about transfers in the last calendar year. In my cleaned sample, the annual transfer rate is 24\%. Thus, the two-year transfer rate could range anywhere between 24\% and 48\%. To be conservative, I assume that the two-year transfer rate equals the one-year transfer rate. My estimate of altruism $\eta$ should be viewed as a lower bound.\footnote{I have experimented with targeted higher transfer rates. This requires stronger altruism, which, in turn, increases homeownership motives, lowering the estimated utility premium $\chi$. Overall, the results are qualitatively similar to my conservative estimate.}

@@

\subsubsection{Household-Level Responses to Income Shocks by Parental Wealth}
As a final external validation, I repeat the exercise in Section \ref{sec:eventstudy} (see Figure \ref{fig:housinggrowthrates}), where we saw that households with wealthy parents do not downsize during unemployment spells. Since the model lacks an explicit unemployment state, households are considered unemployed if their productivity drops to the lowest level for a single period. Otherwise, I replicate the exercise exactly. Figure \ref{fig:housinggrowthrates_model} shows that the model matches both the qualitative and quantitative patterns from the empirical data relatively well: The level of housing consumption falls for households without wealthy parents during ``unemployment,'' while it does not decline for those with rich parents.

@@

\textit{Reducing Purchase Costs:}
Closing-cost programs and first-time-buyer relief---e.g., England’s 2017 stamp duty cut for first-time buyers---directly lower entry costs. I lower the purchase cost ($m_b$) by 2 percentage points, equivalent to a grant of about \$6,000. While homeownership rises, the gains are concentrated among households with wealthy parents, amplifying the parental wealth gradient. Lowering purchase costs has little effect for renters with non-wealthy parents, who remain constrained by LTV limits and sales costs. In contrast, for households with wealthy parents, lower purchase costs reduce the effective price---the main remaining barrier to ownership. This suggests that grants or credits for first-time buyers primarily benefit households with richer parents.

@@

\begin{table}[tb]
	\center\singlespacing	
	\begin{threeparttable}[tb]
		\caption{Altruistic Transfers Decrease the MPC}
		\input{../tabfig/postest/MPCs.tex}
	\end{threeparttable}
	{\begin{footnotesize}\begin{flushleft}\vspace{-0.1in}
		\textit{Notes:} Following \cite{Kaplan2014}, households are liquidity constrained if they have negative liquid wealth $(b_k)$ exceeding 16.5\% of annual earnings. MPCs are calculated from a \$5,000 positive shock to households aged 25-44 in the baseline (with altruism) stationary distribution. ``Future transfers only'' MPCs use the baseline (altruistic) policy functions but set the contemporaneous transfer at the shock date to zero, while leaving expectations about future transfers unchanged. ``Current transfers only'' MPCs combine the parental policy function from the baseline with children’s policy functions solved without altruism; children receive the contemporaneous transfer but behave as if there are no future transfers. \end{flushleft}\end{footnotesize}}			
\end{table}

These results have several implications for the MPC literature. First, model-implied MPCs may be overstated when intra-family transfers are omitted, particularly for households with wealthy parents. Second, a common strategy to match high MPCs is to introduce households with low target wealth---e.g., through impatience \citep{aguiar2024hand}. Altruistic expectations can generate qualitatively similar behavior, as anticipated parental support reduces savings motives. Third, models that generate high MPCs by introducing illiquid assets with large return premia \citep{kaplan2022marginal} may overlook the fact that homeowners with wealthy parents have access to informal insurance, which lowers their effective MPC.

@@

Several simplifying assumptions keep the model tractable. The housing block could be enriched with multiple house sizes, refinancing frictions, and full general equilibrium to capture quality margins and price feedbacks. The dynasty is modeled as a single parent–child pair; in reality, resources are pooled across parents (including remarried/cohabiting partners) and allocated across multiple children. Pooling can raise transfer capacity, while sibling dilution spreads support, making the net strength of the transfer motive ambiguous and heterogeneous across families.  Moreover, parents can also provide non-financial support, such as co-signing mortgages, allowing children to live at home rent-free, or offering informal childcare. These non-financial transfers could substitute for direct transfers. Future work could also study whether illiquidity mitigates intrafamily commitment problems. Finally, the results on illiquidity preferences speak to the high-MPC literature \citep[e.g.,][]{kaplan2022marginal}: models that ignore parental transfers likely overstate average MPCs among liquidity-constrained households, some of whom rely on family support to smooth shocks.

### 2025-10-17 docs/outside_parameters.tex
Re stud (#20)
* some shortening

* old latex is corruped, work on this one in the meantime

* minor changes

* Make error bars thicker

* Better figures

* Small fix

* Update text

* FEDSstuff

* Taking on comments

---------

Co-authored-by: Eirik Eylands Brandsaas <eirik.e.brandsaas@frb.gov>

\begin{tabular}{@{}llll@{}}
    \toprule
    Parameter 			& 		 			& Value  	& Source \\	\midrule
    Period Length		& --				& 2 years   & PSID Frequency \\
    Discount Factor$^\dagger$		& $\beta$			& \parbeta		& Literature (see text)  \\
    Risk Aversion		& $\gamma$			& \pargamma		& Literature (see text)  \\
    Expenditure Share Housing & $\xi$		& \parxi        & Literature (see text) \\
    Rental Price$^\dagger$ 		& $q$				& \parq 		& \cite{Davis2008}  \\
    Deprecation$^\dagger$ 		& $\delta$			& \pardelta		& \cite{Harding2007} \\			
    Risk-free Rate$^\dagger$		& $r^f$ 			& \parrf 	    & Literautre (see text) \\
    Mortgage premium$^\dagger$	& $r^m$ 			& \parrm 	    & Literature (see text) \\
    PMI premium$^\dagger$         & $r^{pmi}$         & \parrpmi      & \cite{goodman2017sixty} \\
    PMI limit           & $\overline{PMI}$   & \parPMIlim     & \cite{goodman2017sixty} \\ 
    House Price          & $p$		& \parprice      & Data (see text) \\
    Max Loan-to-Value   & LTV 			& \parLTV  & Literature/data (see text) \\
    Selling \& Buying Cost		& ($m_s,m_b$) 			& (\parms,\parmb) &  \cite{Yang2009} \\ 
    Rental Size & $h_r$ & \parhr & Normalization \\
    Deterministic Income & $ l_a$				& Fig. \ref{fig:y_a}		& PSID \\
    Productivity Shocks for Kids & $y,\Pi(y'|y)$ & Fig. \ref{fig:y_ia},\ref{fig:Piv}      & PSID \\
    Initial Distribution & $F(x_{53},v_{53}$)	& Fig. \ref{fig:inidistr}		& PSID \\			
    \bottomrule			
\end{tabular}

>>>>

\begin{tabular}{@{}llll@{}}
    \toprule
    Parameter 			& 		 			& Value  	& Source \\	\midrule
    Period Length		& --				& 2 years   & PSID Frequency \\
    Discount Factor$^\dagger$		& $\beta$			& \parbeta		& Literature (see text)  \\
    Risk Aversion		& $\gamma$			& \pargamma		& Literature (see text)  \\
    Expenditure Share Housing & $\xi$		& \parxi        & Literature (see text) \\
    Rental Price$^\dagger$ 		& $q$				& \parq 		& \cite{Davis2008}  \\
    Deprecation$^\dagger$ 		& $\delta$			& \pardelta		& \cite{Harding2007} \\			
    Risk-free Rate$^\dagger$		& $r^f$ 			& \parrf 	    & Literature (see text) \\
    Mortgage premium$^\dagger$	& $r^m$ 			& \parrm 	    & Literature (see text) \\
    PMI premium$^\dagger$         & $r^{pmi}$         & \parrpmi      & \cite{goodman2017sixty} \\
    PMI limit           & $\overline{PMI}$   & \parPMIlim     & \cite{goodman2017sixty} \\ 
    House Price Level   & $p$		& \parprice      & Data (see text) \\
    Max Loan-to-Value   & LTV 			& \parLTV  & Literature/data (see text) \\
    Selling \& Buying Cost		& ($m_s,m_b$) 			& (\parms,\parmb) &  \cite{Yang2009} \\ 
    Rental Size & $h_r$ & \parhr & Normalization \\
    Deterministic Income & $ l_a$				& Fig. \ref{fig:y_a}		& PSID \\
    Productivity Shocks for Kids & $y,\Pi(y'|y)$ & Fig. \ref{fig:y_ia},\ref{fig:Piv}      & PSID \\
    Initial Distribution & $F(x_{53},v_{53}$)	& Fig. \ref{fig:inidistr}		& PSID \\			
    \bottomrule			
\end{tabular}
